---
title: "Debugging in R"
output: html_document
---

```{r}
d <- list.files(here::here("data-raw"), full.names = TRUE,
  pattern = "^The") %>% 
  purrr::map_df(readr::read_csv)
names(d) <- tolower(names(d))
```

# Exercise (browser)

Figure out the bug when you run `get_climates()` below.

Hint: Use `traceback()` to find where it occurs, add `browser()` calls
Hint: Look at the types of the data and think about what the function expects.

```{r}
# Separate, flatten, and trim values in the vector
clean <- function(vec) {
  values <- strsplit(as.character(vec), ",")
  flat_values <- unlist(values)
  trimmed_values <- stringr::str_trim(flat_values)
  trimmed_values
}

# Clean vector and get the unique values
uniquify <- function(vec) {
  clean_values <- clean(vec)
  unique_values <- unique(clean_values)
  unique_values
}

# Read data and get unique climate values
get_climates <- function() {
  planets <- read.csv2(here::here("data-raw/planets.csv"))
  unique_climate <- uniquify(planets$climate)
  unique_climate
}

get_climates()

# This example originally used in Amanda Gadrow's excellent debugging talk at rstudio::conf 2018,
# https://github.com/ajmcoqui/debuggingRStudio/blob/b70a3575a3ff5e7867b05fb5e84568abba426c4b/error_example.R
# and
# https://whattheyforgot.org/debugging-r-code.html
```

Once you have the above working, use `debugonce()` to step into `get_climates()` and look at what the `planets` data frame looks like when it is first read in.

# Exercise (debug)

Use `debugonce()` to step into `hist()`, step through the insanity that is the calculation of the breaks, and figure out what the value of `fuzzybreaks` is.

Hint: it's a ways down on line 88.

```{r}
set.seed(1)
hist(rnorm(100))
```

Use `debugonce()` to step through `ggplot2::ggplot`.

Separately, use `debugonce()` to step into the last internal function used in `ggplot2::ggplot()`. Note that you will have to use `:::`.

```{r, eval=FALSE}
library(ggplot2)
ggplot(mtcars, aes(mpg, cyl)) + geom_point()

debugonce(ggplot2::ggplot) # exercise
ggplot(mtcars, aes(mpg, cyl)) + geom_point()

debugonce(ggplot2:::set_last_plot) # exercise
ggplot(mtcars, aes(mpg, cyl)) + geom_point()
```

# Exercise (trace)

Given the following function, use `trace()` to add a `browser()` statement before the `stop()` and inspect the value of `i`:

Hint: I'll get you started...

```{r, eval=FALSE}
fun <- function() {
  for (i in seq_len(1000)) {
    if (i == 987)
      stop("Bad number!")
  }
}

x <- as.list(body(fun))
x
as.list(x[[2]])
as.list(x[[c(2, 4)]]) # exercise
# keep going...
as.list(x[[c(2, 4, 2)]]) # exercise
as.list(x[[c(2, 4, 2, 3)]]) # exercise
trace(fun, browser, 
  at = list(c(2, 4, # keep going...
    2, 3)  # exercise
    ))

fun()
untrace(fun)
```

# Resources

<https://adv-r.hadley.nz/debugging.html>
<https://whattheyforgot.org/debugging-r-code.html>
