---
title: "Functional programming and iterating with lists and purrr"
output: html_document
---

```{r}
library(dplyr)
library(purrr)
d <- readRDS("data/wip.rds")
wip_dat <- split(d, d$country) %>% # ignore for now
  map(~ .x$percent_wip) # ignore for now
years <- split(d, d$country) %>% # ignore for now
  map(~ .x$year) # ignore for now
```

```{r}
map(wip_dat, mean)
map_dbl(wip_dat, mean)
map_chr(wip_dat, mean)
```

We can pass other arguments to our main function through the `...`:

```{r}
map_dbl(wip_dat, quantile)
map_dbl(wip_dat, quantile, prob = 0.8)
```

Suppose we want to do something more complicated that isn't in an existing function. How do that? We write our own function!

What if we want to calculate the mean women in parliament percentage for each country after dropping the lowest year?

1. Write code that solves the problem for a real object

```{r}
vec <- wip_dat$Canada
(sum(vec) - min(vec)) / (length(vec) - 1)

vec <- wip_dat$`United States`
(sum(vec) - min(vec)) / (length(vec) - 1)
```

2. Wrap the code in `function() {}` and try running it:

```{r}
vec <- wip_dat$Canada
mean_wip <- function() {
  (sum(vec) - min(vec)) / (length(vec) - 1)
}
mean_wip()
```

3. Add the name of the temporary object as an argument. Supply new values to use for each of the arguments.

```{r}
mean_wip <- function(vec) {
  (sum(vec) - min(vec)) / (length(vec) - 1)
}
mean_wip(vec = wip_dat$Canada)
mean_wip(vec = wip_dat$`United States`)
```

4. Change the argument name if desired:

```{r}
mean_wip <- function(x) {
  (sum(x) - min(x)) / (length(x) - 1)
}
mean_wip(wip_dat$`United States`)
```

5. Iterate with your new function:

```{r}
map_dbl(wip_dat, mean_wip)
```

# Exercise

- Calculate the variance of the percentage of women in parliament for each country. Return the result as a numeric vector.
- Write a function that calculates the range (maximum minus minimum value) of percentage of women in parliament for a single country. Apply it to Canada and Finland.
- Apply your range function to all countries and return the result as a numeric vector. As a final step, sort the output from lowest to highest value.

```{r}
map_dbl(wip_dat, var) # exercise

range_wip <- function(x) { # exercise
  max(x) - min(x) # exercise
} # exercise
range_wip(wip_dat$Canada) # exercise
range_wip(wip_dat$Finland) # exercise

map_dbl(wip_dat, range_wip) # exercise
map_dbl(wip_dat, range_wip) %>% sort() # exercise
```

# Anonymous functions

For short function sometimes is not worth naming your function. Instead we can you something called anonymous functions.

As an example, we could rewrite our calculations above that take the mean women in parliament percentage for each country after dropping the lowest year using an anonymous function:

```{r}
map_dbl(wip_dat, function(x) (sum(x) - min(x)) / (length(x) - 1))
```

`function(x) (sum(x) - min(x)) / (length(x) - 1)` exists internally but does not have a name. It is an anonymous function. Usually they should be short, ideally fitting on 1 or maybe 2 lines.

purrr has a shortcut for anonymous functions because they are a very common and useful way of working:

```{r}
map_dbl(wip_dat, ~ (sum(.x) - min(.x)) / (length(.x) - 1))
```

If you are collaborating with people who haven't used purrr, they might find this confusing!

# Iterating with multiple inputs

So far we've only worked with a single input argument in our functional programming. Often that's all you need, but sometimes you need multiple inputs.

purrr contains `map2` variants of its map functions that our specialized for 2 inputs and `pmap` (p = parallel) variants that work with an unlimited number of inputs. 

Note this is not parallel processing on multiple cores, this is parallel in the sense of the `pmin()` or `pmax()` functions.

The `map2()` functions create named inputs of `.x` and `.y` for the first and second arguments. For example, let's fit a linear regression to the women in parliament data over time for each country (we will use an anonymous function):

```{r}
map2(years, wip_dat, ~ lm(.y ~ .x))
map2(years, wip_dat, function(.x, .y) lm(.y ~ .x)) # equivalent
```

If instead we wanted to use the more generic `pmap()`, we could do it like this:

```{r}
pmap(list(years, wip_dat), function(x, y) lm(y ~ x))
```

Notice how you could extend this to have an unlimited number of arguments.

# Exercise

The following is a short function to return the slope from a regression of the percentage of women in parliament over time:

```{r}
# x = years and y = wip_dat
get_slope <- function(x, y) {
  m <- lm(y ~ x)
  coef(m)[[2]]
}
```

Use `map2_dbl()` to return the slope for each country. Try it 3 ways:

1. The way written below
2. Via the `~` shortcut to an anonymous function.
3. Via the standard R anonymous function syntax.

Why might you pick one over another?

Try passing the results through `sort()` and optionally `rev()` to find the most positive and negative slopes.

```{r}
map2_dbl(years, wip_dat, get_slope)
map2_dbl(years, wip_dat, ~ get_slope(.x, .y)) # exercise
map2_dbl(years, wip_dat, function(.x, .y) get_slope(.x, .y)) # exercise

map2_dbl(years, wip_dat, get_slope) %>% sort() # exercise
map2_dbl(years, wip_dat, get_slope) %>% sort() %>% rev() # exercise
```

# Data frames are lists! And tricks for extra variables

Remember that data frames are lists so they can be passed as input to `pmap()`. Since data manipulation often happens in a data frame, this is a useful thing to remember.

However, we often end up in a situation like the following:

```{r}
df <- data.frame(
  x = 1:3 + 0.1,
  y = 3:1 - 0.1,
  z = letters[1:3]
)
```

Why doesn't the following work?

```{r, error=TRUE}
plus <- function(x, y) x + y
pmap_dbl(df, plus)
```

How can we fix it? There are at least 2 ways. Think about it for a minute.

Hint: one of the solutions involves `...`.

```{r}
plus2 <- function(x, y, ...) x + y  # exercise
pmap_dbl(df, plus2)

plus3 <- function(x, y, z) x + y  # exercise
pmap_dbl(df, plus3)
```

# Exercise

Let's put this all together and try the following:

1. Make a function to calculate the slope of the percentage of women in parliament for each country but only for the years 2010 and after.
2. Make the 2010 an argument named `cutoff` and try passing it via the `...` with your `map()` function.
- Rerun your calculation for all years 2005 and after using your new function.

```{r}
get_slope2 <- function(yrs, wip, cutoff = 2010) { # exercise
  wip <- wip[yrs >= cutoff] # exercise
  yrs <- yrs[yrs >= cutoff] # exercise
  m <- lm(wip ~ yrs) # exercise
  coef(m)[[2]] # exercise
} # exercise
map2_dbl(years, wip_dat, get_slope2) %>% sort() # exercise
map2_dbl(years, wip_dat, get_slope2, cutoff = 2005) %>% sort() # exercise
```

# Bonus question

Was that too easy? If so, try the following next-level purrr exercise in one set of pipes:

- Using your last function, use `map()` to also iterate over year thresholds of `seq(1990, 2010, 5)`.
- Use `map()` to `sort()` the output from highest to lowest value (i.e. also use `rev()`).
- Use `[`() to extract only the top 5 countries for each cut off.

Yes, there are many other tidyverse ways to do this, but try to do it with purrr and base R functions as an exercise.

```{r}
map(seq(1990, 2010, 5), function(z) # exercise
  map2_dbl(years, wip_dat, get_slope2, cutoff = z)) %>% # exercise
  map(sort) %>%  # exercise
  map(rev) %>%  # exercise
  map(`[`, i = 1:5) # exercise
```

