---
title: "Functional programming and iterating with lists and purrr"
output: html_document
---

# Goals

- Be able to explain what "apply" functions are and why they are useful.
- Be able to apply a basic function to a list or vector with `purrr::map()`.
- Understand what anonymous functions are and where they are useful.
- Gain an overview of the `purrr::map()` output variants and what they do. `purrr::map_dbl()` etc.
- Become familiar with other multi-input purrr functions: `purrr::map2()`,`purrr::pmap()`.
- Become aware of the breadth of other functions available in purrr (e.g. `purrr::walk()`, `purrr::pluck()`)

# An introduction to iterating with purrr

First read in the following data set:

```{r}
library(dplyr)
library(purrr)
d <- readRDS(here::here("data/wip.rds"))
wip_dat <- split(d, d$country) %>% # ignore for now
  map(~ .x$percent_wip) # ignore for now
years <- split(d, d$country) %>% # ignore for now
  map(~ .x$year) # ignore for now
```

These data represent the percentage of women in parliament for various countries over time:

```{r}
head(d)
```

I have converted the data frame into 2 lists split by country for some exercises:

```{r}
wip_dat[1]
wip_dat[[1]]
years[1]
```

If we wanted to calculate the mean for certain country, we could do it like this:

```{r}
mean(wip_dat[[1]])
```

`purrr::map()` let us iterate over all the elements of the last and apply our function:

```{r results="hide"}
map(wip_dat, mean)
```

We are now "functional programming" because one of the arguments to the function `map()` is another function!

`purrr::map()` returns a list but there are other "type stable" functions within purrr to return different object types. For example:

```{r results="hide"}
map_dbl(wip_dat, mean)
map_chr(wip_dat, mean)
```

We can pass other arguments to our main function through the `...`:

```{r results="hide"}
map(wip_dat, quantile)
map(wip_dat, quantile, prob = 0.8)
```

# Functional programming with our own functions

Suppose we want to do something more complicated that isn't in an existing function. How do that? We write our own function!

What if we want to calculate the mean women in parliament percentage for each country after dropping the lowest year?

1. Write code that solves the problem for a real object

```{r}
vec <- wip_dat$Canada
(sum(vec) - min(vec)) / (length(vec) - 1)

vec <- wip_dat$`United States`
(sum(vec) - min(vec)) / (length(vec) - 1)
```

2. Wrap the code in `function() {}` and try running it:

```{r}
vec <- wip_dat$Canada
mean_wip <- function() {
  (sum(vec) - min(vec)) / (length(vec) - 1)
}
mean_wip()
```

3. Add the name of the temporary object as an argument. Supply new values to use for each of the arguments.

```{r}
mean_wip <- function(vec) {
  (sum(vec) - min(vec)) / (length(vec) - 1)
}
mean_wip(vec = wip_dat$Canada)
mean_wip(vec = wip_dat$`United States`)
```

4. Change the argument name if desired:

```{r}
mean_wip <- function(x) {
  (sum(x) - min(x)) / (length(x) - 1)
}
mean_wip(wip_dat$`United States`)
```

5. Iterate with your new function:

```{r results="hide"}
map_dbl(wip_dat, mean_wip)
```

# Exercise

- Calculate the variance of the percentage of women in parliament for each country. Return the result as a numeric vector.
- Write a function that calculates the range (maximum minus minimum value) of percentage of women in parliament for a single country. Apply it to Canada and Finland.
- Apply your range function to all countries and return the result as a numeric vector. As a final step, sort the output from lowest to highest value.

```{r results="hide"}
map_dbl(wip_dat, var) # exercise

range_wip <- function(x) { # exercise
  max(x) - min(x) # exercise
} # exercise
range_wip(wip_dat$Canada) # exercise
range_wip(wip_dat$Finland) # exercise

map_dbl(wip_dat, range_wip) # exercise
map_dbl(wip_dat, range_wip) %>% sort() # exercise
```

# Exercise (reading multiple data files with purrr)

Here's a character vector with 3 .csv files to be read in.

```{r}
f <- here::here("data-raw",
  c("The_Fellowship_Of_The_Ring.csv", 
    "The_Return_Of_The_King.csv", 
    "The_Two_Towers.csv"
  ))
```

Notice how I used `here::here()` and what that is doing here.

Sometimes, `list.files()` is useful for creating a character vector of filenames.

For example:

```{r}
list.files(here::here("data-raw"), full.names = TRUE, 
  pattern = "^The")
```

Use `map_df()` to read all three files in and bind them into a data frame in one line of code. Try with `readr::read_csv` and try with `read.csv` while passing the option `stringsAsFactors = FALSE`.

```{r}
map_df(f, readr::read_csv) # exercise
map_df(f, read.csv, stringsAsFactors = FALSE) # exercise
```

# Anonymous functions

For short functions, sometimes it's not worth naming your function. Instead we can use something called anonymous functions.

As an example, we could rewrite our calculations above that take the mean women in parliament percentage for each country after dropping the lowest year using an anonymous function:

```{r results="hide"}
map_dbl(wip_dat, function(x) (sum(x) - min(x)) / (length(x) - 1))
```

`function(x) (sum(x) - min(x)) / (length(x) - 1)` exists internally but does not have a name. It is an anonymous function. Usually these should be short, ideally fitting on 1 or maybe 2 lines.

purrr has a shortcut for anonymous functions because they are a very common and useful way of working:

```{r results="hide"}
map_dbl(wip_dat, ~ (sum(.x) - min(.x)) / (length(.x) - 1))
```

If you are collaborating with people who haven't used purrr, they might find this confusing!

# Iterating with multiple inputs

So far we've only worked with a single input argument. Sometimes you need multiple inputs.

purrr contains `map2` variants of its map functions that our specialized for 2 inputs and `pmap` (p = parallel) variants that work with an unlimited number of inputs. Note this is not parallel processing on multiple cores, this is parallel in the sense of iterating over multiple arguments in parallel.

The `map2()` functions create named inputs of `.x` and `.y` for the first and second arguments. For example, let's fit a linear regression to the women in parliament data over time for each country (we will use an anonymous function):

```{r results="hide"}
map2(years, wip_dat, ~ lm(.y ~ .x))
```

If instead we wanted to use the more generic `pmap()`, we could do it like this:

```{r results="hide"}
pmap(list(years, wip_dat), function(x, y) lm(y ~ x))
```

Notice how you could extend this to have an unlimited number of arguments.

# Exercise

The following is a short function to return the slope from a regression of the percentage of women in parliament over time:

```{r}
# x = years and y = wip_dat
get_slope <- function(x, y) {
  m <- lm(y ~ x)
  coef(m)[[2]]
}
```

Use `map2_dbl()` to return the slope for each country. Try it 3 ways:

1. The way written below
2. Via the `~` shortcut to an anonymous function.
3. Via the standard R anonymous function syntax.

Why might you pick one over another?

Try passing the results through `sort()` and optionally `rev()` to find the most positive and negative slopes.

```{r results="hide"}
map2_dbl(years, wip_dat, get_slope)
map2_dbl(years, wip_dat, ~ get_slope(.x, .y)) # exercise
map2_dbl(years, wip_dat, function(.x, .y) get_slope(.x, .y)) # exercise

map2_dbl(years, wip_dat, get_slope) %>% sort() # exercise
map2_dbl(years, wip_dat, get_slope) %>% sort() %>% rev() # exercise
```

# Data frames are lists! And tricks for extra variables

Remember that data frames are lists so they can be passed as input to `pmap()`. Since data manipulation often happens in a data frame, this is a useful thing to remember.

However, we often end up in a situation like the following:

```{r}
df <- data.frame(
  x = 1:3 + 0.1,
  y = 3:1 - 0.1,
  z = letters[1:3]
)
```

Why doesn't the following work?

```{r, error=TRUE}
plus <- function(x, y) x + y
pmap_dbl(df, plus)
```

How can we fix it? There are at least 2 ways. Think about it for a minute.

Hint: one of the solutions involves `...`.

```{r}
plus2 <- function(x, y, ...) x + y  # exercise
pmap_dbl(df, plus2)

plus3 <- function(x, y, z) x + y  # exercise
pmap_dbl(df, plus3)
```

# Exercise

Let's put this all together and try the following:

1. Make a function to calculate the slope of the percentage of women in parliament for each country but only for the years 2010 and after.
2. Make the 2010 an argument named `cutoff` and try passing it via the `...` with your `map()` function.
- Rerun your calculation for all years 2005 and after using your new function.

```{r results="hide"}
get_slope2 <- function(yrs, wip, cutoff = 2010) { # exercise
  wip <- wip[yrs >= cutoff] # exercise
  yrs <- yrs[yrs >= cutoff] # exercise
  m <- lm(wip ~ yrs) # exercise
  coef(m)[[2]] # exercise
} # exercise
map2_dbl(years, wip_dat, get_slope2) %>% sort() # exercise
map2_dbl(years, wip_dat, get_slope2, cutoff = 2005) %>% sort() # exercise
```

# Bonus question

Was that too easy? If so, try the following next-level purrr exercise in one set of pipes:

- Using your last function, use `map()` to also iterate over year thresholds of `seq(1990, 2010, 5)`.
- Use `map()` to `sort()` the output from highest to lowest value (i.e. also use `rev()`).
- Use `[`() to extract only the top 5 countries for each cut off.

Yes, there are many other tidyverse ways to do this, but try to do it with purrr and base R functions as an exercise.

```{r}
map(seq(1990, 2010, 5), function(z) # exercise
  map2_dbl(years, wip_dat, get_slope2, cutoff = z)) %>% # exercise
  map(sort) %>%  # exercise
  map(rev) %>%  # exercise
  map(`[`, i = 1:5) # exercise
```

# Diving deeper into purrr

purrr has a lot of helpful functions for working with lists beyond the `map()` family of functions.

Take 2 minutes to skim through the help for purrr:

```{r}
help(package = "purrr")
```

and try to find 3-5 interesting functions. Think about when you might use those and how they would be helpful.

Take another couple minutes to discuss what you found with your partner.

If you aren't sure where to start, take a look at `purrr::walk()` and `purrr::pluck()`.
