---
title: "Bootstrapping in parallel with the future package"
output: html_document
---

# Goals:

- Learn a variety of strategies for integrating future with dplyr and pipes.
- Practice benchmarking parallel vs. sequential code.
- Become aware of the 'furrr' package for mixing 'purrr' and 'future'.
- Gain experience with a more complicated and realistic application of parallel processing in R.

# The data

Let's load the packages we will use:

```{r}
library(future)
library(tictoc)
library(dplyr)
library(ggplot2)
```

Set up our parallel processing plan:

```{r}
future::plan(multisession)
```

We will work with trawl survey data for Pacific cod in Hecate Strait:

```{r}
d <- readRDS(here::here("data/pcod-syn-hs.rds"))
glimpse(d)
ggplot(d, aes(longitude, latitude, size = density_kgpm2)) +
  geom_point() +
  scale_size_area() +
  facet_wrap(~year)
```

# Functions for bootstrapping a relative biomass index

Let's set up a series of small functions so we can calculate a bootstrap estimate of a groundfish trawl survey biomass index each year.

First, a function that takes our data frame and calculates the biomass "stratified" by the grouping codes (depth strata):

```{r}
calculate_biomass <- function(dat, i = seq_len(nrow(dat))) {
  dat[i, ] %>% group_by(year, grouping_code) %>%
    summarise(density = mean(density_kgpm2)) %>%
    group_by(year) %>%
    summarise(biomass = sum(density * 2 * 2)) %>%
    pull(biomass)
}
```

Study that function and make sure you understand what it is doing. Note that we have set it up with an argument named `i`, which allows us to run our function on some subset of the data frame. This will be important so that we can use the `boot::boot()` function for bootstrapping. We have set it up so that it defaults to during the calculation on all rows.

Let's try out our function:

```{r}
calculate_biomass(d)
calculate_biomass(d, i = 1:200)
```

Now let's create a function that takes our data for a single year and applies a bootstrap stratified by the grouping codes/depth strata:

```{r}
bootstrap_one_year <- function(x, reps) {
  b <- boot::boot(x, statistic = calculate_biomass, strata = x$grouping_code, 
    R = reps)
  suppressWarnings(bci <- boot::boot.ci(b, type = "perc"))
  tibble::tibble(
    survey = x$survey[1],
    year = x$year[1],
    est = mean(b$t),
    lwr = bci$percent[[4]],
    upr = bci$percent[[5]],
    cv = sd(b$t)/mean(b$t))
}
```

Study that function make sure you understand what it is doing.

Let's try that function on a single year and for 50 bootstrap iterations:

```{r}
dplyr::filter(d, year == 2019) %>%
  bootstrap_one_year(reps = 50)
```

Now we need some way to iterate over the various years. Here's one way we can do it sequentially:

```{r}
boot_biomass <- function(dat, reps = 500) {
  dat %>% dplyr::group_split(year) %>% 
    purrr::map_dfr(bootstrap_one_year, reps = reps)
}
b <- boot_biomass(d, reps = 10)
ggplot(b, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_ribbon(alpha = 0.4) +
  geom_line() +
  facet_wrap(~survey)
```

This makes use of the `dplyr::group_split()` function which is a tidyverse version of `base::split()`.

# Bootstrapping in parallel

There are various ways we could run this in parallel with the future package. Let's start with a relatively verbose way of doing it by working with the future package directly. First, as an intermediate step, let's rewrite the above function to use `purrr::map()` and `purrr::dfr()` with anonymous functions:

```{r}
boot_biomass2 <- function(dat, reps = 500) {
  dat %>% dplyr::group_split(year) %>% 
    purrr::map(function(.x) bootstrap_one_year(.x, reps = reps)) %>% 
    purrr::map_dfr(function(.x) .x)
}
```

Remember that purrr includes a shortcut for this (`~` replaces `function(.x)` and the passed value is available as `.x`). Try converting to the shortcut syntax:

```{r}
boot_biomass3 <- function(dat, reps = 500) {
  dat %>% dplyr::group_split(year) %>% 
    purrr::map(~bootstrap_one_year(.x, reps = reps)) %>% # exercise
    purrr::map_dfr(~.x) # exercise
}
```

Let's prove to ourselves that these all do the same thing:

```{r}
set.seed(1)
boot_biomass(d, reps = 50)
set.seed(1)
boot_biomass2(d, reps = 50)
set.seed(1)
boot_biomass3(d, reps = 50)
```

Now, we can do the same in parallel by using the `future::future()` and `future::value()` functions:

```{r}
boot_biomass_purrr_future <- function(dat, reps = 500) {
  dat %>% dplyr::group_split(year) %>% 
    purrr::map(~ future(bootstrap_one_year(.x, reps = reps))) %>%
    purrr::map_dfr(~ value(.x))
}
```

It's a little bit ugly but it works:

```{r}
boot_biomass_purrr_future(d, reps = 50)
```

Let's look at two slightly cleaner ways of doing this. First is with the 'furrr' package, which is a drop-in replacement for many of the 'purrr' functions using the future package in the background for parallel processing:

```{r}
boot_biomass_furrr <- function(dat, reps = 500) {
  dat %>% dplyr::group_split(year) %>% 
    furrr::future_map_dfr(bootstrap_one_year, reps = reps)
}
boot_biomass_furrr(d, reps = 50)
```

And the second is to use the 'future.apply' package. This package has replacements for many apply functions. For example, `future.apply::future_lapply()` is a replacement for the `lapply()` function. Remember that `purrr::map()` is analogous to `lapply()`: it applies a function to multiple elements and returns a list.

```{r}
boot_biomass_future.apply <- function(dat, reps = 500) {
  dat %>% dplyr::group_split(year) %>% 
    future.apply::future_lapply(bootstrap_one_year, reps = reps) %>% 
    dplyr::bind_rows()
}
boot_biomass_future.apply(d, reps = 50)
```

How do we know that our parallel functions actually saved us time? One quick way to check is with `system.time()` or the 'tictoc' package:

```{r}
tic("sequential")
boot_biomass(d, reps = 500)
toc()

tic("purrr")
boot_biomass_purrr_future(d, reps = 500)
toc()

tic("furrr")
boot_biomass_furrr(d, reps = 500)
toc()

tic("future.apply")
boot_biomass_future.apply(d, reps = 500)
toc()
```

A more rigourous test would be with benchmarking:

```{r}
b <- bench::mark(
  sequential = boot_biomass(d, reps = 24),
  future.apply = boot_biomass_future.apply(d, reps = 24),
  purrr_futur = boot_biomass_purrr_future(d, reps = 24),
  furrr = boot_biomass_furrr(d, reps = 24),
  iterations = 10, check = FALSE)
print(b)
plot(b)
```

Why did we set the `iterations` and `check` argument explicitly? What would be an alternative to setting `check = FALSE`?

# Exercise

Here is the data for Hecate Strait, Queen Charlotte Sound, and West Coast Vancouver Island:

```{r}
d_multiple <- readRDS(here::here("data", "pcod-syn-hs-qcs-wcvi.rds"))
glimpse(d_multiple)
ggplot(d_multiple, aes(longitude, latitude, size = density_kgpm2, colour = survey)) +
  geom_point() +
  scale_size_area() +
  facet_wrap(~year)
```

Try extending our above function (`boot_biomass()`) so that it also works with multiple surveys (start by copying and pasting the original function below). Then pick one of the above approaches and use it to iterate over years and surveys in parallel in a separate function.

Check to make sure your parallel version is faster.

```{r}
boot_biomass2 <- function(dat, reps = 200) { # exercise
  dat %>% dplyr::group_split(year, survey) %>% # exercise
    purrr::map_dfr(bootstrap_one_year, reps = reps) # exercise
} # exercise

boot_biomass_future.apply2 <- function(dat, reps = 200) { # exercise
  dat %>% dplyr::group_split(year, survey) %>% # exercise
    future.apply::future_lapply(bootstrap_one_year, reps = reps) %>% # exercise
    dplyr::bind_rows() # exercise
}

tic() # exercise
b <- boot_biomass2(d_multiple) # exercise
toc() # exercise
b # exercise

tic() # exercise
b <- boot_biomass_future.apply2(d_multiple) # exercise
toc() # exercise
b # exercise
```

Assuming you saved your output to an object named `b`, you can plot it with the following code:

```{r}
ggplot(b, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_ribbon(alpha = 0.4) + 
  geom_line() +
  facet_wrap(~survey)
```
