---
title: "Benchmarking"
output: html_document
---

# Goals

- Understand what benchmarking is and why it's important.
- Gain familiarity with using the bench package.
- Practice benchmarking some R functions.

# The bench package

There are multiple benchmarking packages in R. We're going to work with the bench package, which has the cutely named main function `mark()` so you can call it as `bench::mark()`.

The basic format is like this:

```{r}
dat <- data.frame(x = runif(100, 1, 1000), y = runif(10, 1, 1000))
b <- bench::mark(
  dat[dat$x > 500, ],
  dplyr::filter(dat, x > 500),
  subset(dat, x > 500),
  check = FALSE
)
b
plot(b)
```

Read the help file for `?bench::mark()` to look at the various options.

Answer the following:

1. What does the `check` argument do? When might you want to set it to `FALSE` vs. `TRUE`?
2. What does `relative` do?
2. How do `min_time`, `iterations`, `min_iterations`, and `max_iterations` interact?

# Exercises

We'll use benchmarking to test a number of suggestions for speeding up R code.

## Look for existing solutions

Benchmark `lm()` vs. `lm.fit()` for fitting a model of `y` vs. `x`. While you're at it, skim the help file for `lm.fit()`. You should usually be using `lm()` even if it's slower!

```{r}
x <- rnorm(1e6)
y <- 0.5 + 0.2 * x + rnorm(length(x), sd = 0.1)
model_matrix <- matrix(cbind(rep(1, length(x)), x), ncol = 2)
head(model_matrix)

b <- bench::mark(
  m <- lm(y ~ x), # exercise
  m2 <- lm.fit(model_matrix, y), # exercise
  check = FALSE
)
b
plot(b)
```

## Do as little as possible

Benchmark `rowMeans()` vs. `.rowMeans()`:

```{r}
set.seed(123)
df <- cbind(x1 = rnorm(1e3), x2 = rnorm(1e3))
head(df)

b <- bench::mark(
  rowMeans(df), # exercise
  .rowMeans(df, m = 1e3, n = 2) # exercise
)
b
plot(b)
```

## Vectorize

Benchmark `remove_NAs1()` vs. `remove_NAs2()`:

```{r}
x <- rnorm(1e4)
x[c(2, 200, 400)] <- NA

remove_NAs1 <- function(x) {
  for (i in seq_along(x)) {
    if (is.na(x[i])) x[i] <- 0
  }
  x
}

remove_NAs2 <- function(x) {
  x[is.na(x)] <- 0
  x
}


b <- bench::mark(
  remove_NAs1(x), # exercise
  remove_NAs2(x) # exercise
)
b
plot(b)
```

## Avoid copies

Benchmark `my_cumsum1()` vs. `my_cumsum2()`. While you're at it, also compare these to `base::cumsum()`. Which of the above rules does `cumsum()` fall into? Why is it faster?

```{r}
x <- rnorm(1e4)

my_cumsum1 <- function(x) {
  sum <- x[1]
  for (i in seq(2, length(x))) {
    sum[i] <- sum[i - 1] + x[i]
  }
  sum
}

my_cumsum2 <- function(x) {
  sum <- numeric(length(x))
  sum[1] <- x[1]
  for (i in seq(2, length(x))) {
    sum[i] <- sum[i - 1] + x[i]
  }
  sum
}

b <- bench::mark(
  my_cumsum1(x), # exercise
  my_cumsum2(x), # exercise
  cumsum(x) # exercise
)
b
plot(b)
```

## Parallelize

Stay tuned...

## Re-write in C++

Stay tuned...

# More information

- <https://adv-r.hadley.nz/perf-improve.html>
- <http://www.burns-stat.com/pages/Tutor/R_inferno.pdf>
