---
title: "Profiling R code"
output: html_document
---

# Goals

- Understand what code profiling is and why it's important.
- Gain familiarity with using the profvis package.
- Practice profiling some R functions.

# Never forget...

> Programmers waste enormous amounts of time thinking about, or worrying about, the speed of noncritical parts of their programs, and these attempts at efficiency actually have a strong negative impact when debugging and maintenance are considered.

> — Donald Knuth.

# The profvis package

There are a number of profiling options in R. Here we will work with the profvis package. The initial examples below come from the package itself.

Let's start with a simple example from Hadley's Advanced R book:

```{r, eval=FALSE}
source(here::here("performance", "profiling-example.R"))
profvis::profvis(f())
```

Inspect the output and see if you can figure out what everything is telling you.

- Try double-clicking on elements of the "flame graph". Note this only works if the code is in a separate .R file (not in an .Rmd).
- Why is so much time spent on line 11?

Try the following example from the package documentation:

```{r, eval=FALSE}
library(profvis)
profvis({
  data(diamonds, package = "ggplot2")
  plot(price ~ carat, data = diamonds)
  m <- lm(price ~ carat, data = diamonds)
  abline(m, col = "red")
})
```

- Where is the majority of the time spent?

# Exercise

In pairs, take a short chunk of code you have written that takes more than a second or two to run and try profiling it. What can you identify as the bottlenecks? Can you think of obvious solutions to those bottlenecks or are they inevitable?
