---
title: "Rcpp introduction"
output: html_document
---

The most basic way we can evaluate C++ code is via `evalCpp()`. This generally isn't too useful except for testing:

```{r}
library(Rcpp)

evalCpp("1 + 1")
evalCpp("Rcpp::rnorm(1, 0, 1)")
```

The next most basic way we can create a C++ function is via the `cppFunction()` function. This is useful for very simple functions where it's not worth the effort to create a separate file.

For example, take the following R function:

```{r}
one_r <- function() {
  1L
}
one_r()
```

Remember that `1L` explicitly means an integer.

We can create the same function with Rcpp:

```{r}
cppFunction('int one_cpp() {
  return 1;
}')
one_cpp()
```

Note the differences:

- we have to declare the type of the output (`int`)
- we have to explicitly use a `return` statement
- we have to end each line of code with a `;`
- we don't use the `function()` R syntax
- 1 is implicitly an integer in C++ (`1.0` or `1.` would be a real number)

We can extend the above example to take multiple arguments:

```{r}
cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')

add
add(1, 2, 3)
```

We can print out values or error messages using `Rcpp::Rcout` and `Rcpp::Rcerr`. Take a look at the example in `rcpp/print.cpp`:

```{r}
readLines("rcpp/print.cpp")
```


```{r}
sourceCpp("rcpp/print.cpp")
rcpp_rcout(c(1, 2))
```




We can make use of R functions---even functions we write in our global environment---inside Rcpp. To do that we can capture our function from our global environment and apply it:

```
  Environment my_env = Environment::global_env();
  Function sumR = my_env["sumR"];
  return sumR(x);
```

```{r}
sumR <- function(x) {
  sum(x)
}
sumR(c(1, 2, 3))

sourceCpp("rcpp/global-R-function.cpp")
sumC(c(1, 2, 3))
```

As proof, let's modify our `sumR()` function to do something ridiculous and run our C++ function again:

```{r}
sumR <- function(x) {
  sum(x) + 9999
}
sumC(c(1, 2, 3))
```

Of course, our C++ function won't be any faster than the R one, but sometimes this can get you out of having to code something that is tricky but fast in R.

# Exercise

1. What is wrong with the following C++ function? Can you fix it?

(Don't be scared by the verbose C++ error messages!)

```{r, eval=FALSE}
cppFunction('
int add(int x, int y) {
  z = x + y;
  return z;
}
')
```

2. What is wrong with the following C++ function? Can you fix it?

```{r, eval=FALSE}
cppFunction('
double f1(NumericVector x) {
  int n = x.size();
  double y = 0.0;

  for (int i = 0; i < n; ++i) {
    y += x[i] / n;
  }
  return y;
}
')
```

3. What is the line `y += x[i] / n;` doing?

4. What is the above function doing? What R function does this resemble?

5. Can you implement `cumprod()` in Rcpp? I've started below for you:

```{r, eval=FALSE}
cppFunction('
NumericVector cumprod_cpp(NumericVector x) {
  int n = x.size();
  NumericVector y(n);

  // Do the calculations here:
  // Note that there are multiple ways to do this.
  // You may not use the same number of lines.
  
  for (int i = 0; i < n; ++i) { // exercise
    if (i == 0) { // exercise
      y[i] = x[i]; // exercise
    } else { // exercise
      y[i] = y[i-1] * x[i]; // exercise
    } // exercise
  } // exercise
  
  return y;
}
')
cumprod_cpp(c(1, 2, 3, 4))
cumprod(c(1, 2, 3, 4))
```

# Exercise

I've written a Ricker/logistic population dynamic simulation in `rcpp/ricker.R`. Take that code and translate it into C++ for use with Rcpp. Name your new function `ricker_rcpp()`.

Start by verifying that you get the same answer out of both functions if you start with the same random number generator seed.

Benchmark your Rcpp version against R version. Start with a time series of length 100. Then try it with a time series of length 1e4. How do they compare and why?

```{r}
source(here::here("rcpp/ricker.R"))
sourceCpp(here::here("rcpp/ricker-complete.cpp")) # exercise

set.seed(42) # exercise
ricker_r(10) # exercise
set.seed(42) # exercise
ricker_cpp(10) # exercise

bench::mark( # exercise
  ricker_r(100), # exercise
  ricker_cpp(100), # exercise
  check = FALSE # exercise
) # exercise
bench::mark( # exercise
  ricker_r(1e4), # exercise
  ricker_cpp(1e4), # exercise
  check = FALSE # exercise
) # exercise
```

Bonus step: If you got this far (great!), try using `purrr::pmap()` or `purrr:map_dfr()` to run your `ricker_cpp()` function over a series of r and K values for 10 iterations each. Plot the timeseries that you generate.

Hint 1: Start with `expand.grid()` or `tidyr::crossing()`.
Hint 2: `pmap()` takes a list as input and a data frame is also a list.
Hint 3: Except for optional linebreaks, you can do this whole bonus exercise in 3 lines of code!

```{r}
df <- expand.grid(i = 1:10, r = c(0.8, 1.3, 1.9), K = c(5, 15)) # exercise

out <- purrr::pmap_dfr(df, function(r, K, i) # exercise
  data.frame(time = 1:20, i = i, r = r, K = K, # exercise
    y = ricker_cpp(r, K, n = 20))) # exercise

library(ggplot2) # exercise
ggplot(out, aes(time, y, group = i)) +  # exercise
  geom_line() + # exercise
  facet_grid(r ~ K) # exercise
```

# More reading

- <https://adv-r.hadley.nz/rcpp.html>
- <https://thecoatlessprofessor.com/programming/cpp/unofficial-rcpp-api-documentation/>
